The truncated KdV framework for modeling anomalous waves induced by abrupt depth changes

### 2019-07-18 WavesData-2019/LatexNotes/notes.tex
NM

% Basic physical parameters and scales
\newcommand{\freqp}{f_p}
\newcommand{\amp}{a}
\newcommand{\ampscale}{A}
\newcommand{\etastd}{\eta_{\text std}}
% Parameters that change crossing the ADC
\newcommand{\depth}{d}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\dupdn}{\depth_{\pm}}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamdn}{\lam_{+}}
\newcommand{\lamupdn}{\lam_{\pm}}
\newcommand{\lamfac}{N}
\newcommand{\drat}{D}
\newcommand{\dratdn}{\drat_+}
\newcommand{\dratupdn}{\drat_{\pm}}
% Statistical quantities
\newcommand{\skw}{\text{skew}}
\newcommand{\var}{\text{var}}
\newcommand{\En}{\mathcal{E}}
\newcommand{\Mo}{\mathcal{M}}
% Dimensionless parameters
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
% Hamiltonian and Gibbs stuff
\newcommand{\uhat}{\hat{u}}
\newcommand{\Ham}{H}
\newcommand{\Hup}{\Ham^{-}}
\newcommand{\Hdn}{\Ham^{+}}
\newcommand{\Hupdn}{\Ham^{\pm}}
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gup}{\Gibbs^{-}}
\newcommand{\Gdn}{\Gibbs^{+}}
\newcommand{\Gupdn}{\Gibbs^{\pm}}
\newcommand{\thup}{\theta^{-}}
\newcommand{\thdn}{\theta^{+}}
\newcommand{\thupdn}{\theta^{\pm}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}
\newcommand{\meanupdn}[1]{\mean{#1}_{\pm}}

@@

We will essential use the original normalization in \eqref{dimlessKdV1}, except with the slight modification that $\eta = \sqrt{\pi} \etastd u$, so that the energy of $u$ is normalized to unity. This choice will be consistent with our PNAS paper and the J Stat Phys paper of Majda and Qi.
Thus the dimensionless variables are
\begin{align}
&u = \eta/(\sqrt{\pi} \etastd)
&&\mbox{\em dimensionless surface displacement} \\
&\tilde{x} = \frac{\pi \xi}{\lamfac \lam} = \frac{\pi (x-ct)}{\lamfac \lam}
&&\mbox{\em dimensionless position in moving frame} \\
& \tilde{t} = t \freqp / \lamfac
&&\mbox{\em dimensionless time}
\end{align}
The explanation will be something like: instead of the most conventional choice, we scale displacement on $\sqrt{\pi} \etastd$ so that energy is normalize to unity. This choice will be the most convenient for the Hamiltonian framework that we will rely on. Maybe we will have to scale time differently too to get the desired powers of $\drat$, but that will make no difference for now.
Thus, the dimensionless KdV is given by
\begin{align}
\label{dimlessKdV2}
&u_t + C_3 \drat^{-1} \, u u_x + C_2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi^{3/2} \epsup  \\
&C_2 = \frac{\pi^3 \delup^2}{6 \lamfac^2} 
\end{align}
When we draw a comparison to the PNAS paper, it will be something like $\epsup = E^{1/2}L^{-3/2}$ and $\delup^2 = L^{-3}$, etc.

@@

The statistical matching condition is
\begin{equation}
\meanup{\Hdn} = \meandn{\Hdn}
\end{equation}
An experimental observation is that the incoming skewness is small, thus
\begin{equation}
\meanup{H_3} \approx 0
\end{equation}
which immediately gives
\begin{equation}
C_3 \dratdn^{-1} \meandn{H_3} = C_2 \dratdn \left( \meandn{H_2} - \meanup{H_2} \right)
\end{equation}
Simplifying gives
\begin{equation}
\label{H3H2}
\meandn{H_3} = \frac{C_2}{C_3} \dratdn^2 \,\, \Delta \mean{H_2}
\end{equation}
where $\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2}$  indicates the difference between the upstream and downstream values.
To convert to dimensional, i.e.~experimental, values we note
\begin{align}
&\mean{u^3} =  \frac{\mean{\eta^3}} {\pi^{3/2} \mean{\eta^2}^{3/2}} = \pi^{-3/2} \skw(\eta) \\
&\mean{u_x^2} = \pi^{-1} \left(\frac{\lamupdn}{\etastd} \right)^2 \mean{\eta_x^2} 
= \pi^{-1} \left(\frac{\lamupdn}{\etastd} \right)^2 \var(\eta_x)
\end{align}
Therefore
\begin{align}
& \meandn{H_3} = \frac{\pi}{3} \mean{u^3} = 
\frac{1}{3} \pi^{-1/2} \skw_+(\eta) \\
& \meanupdn{H_2} = \pi \meanupdn{u_x^2} = 
\left(\frac{\lamupdn}{\etastd} \right)^2 \var(\eta_x)
\end{align}
The second equation implies
\begin{equation}
\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2} = 
\etastd^{-2} \left( \lamdn^2 \var_+(\eta_x) - \lamup^2 \var_{-}(\eta_x) \right) =
\etastd^{-2} \Delta \left( \lam^2 \var(\eta_x) \right)
\end{equation}
Then \eqref{H3H2} gives the relationship
\begin{equation}
\frac{\skw(\eta)} {\Delta \left( \lam^2 \var(\eta_x) \right) } =
\frac{3 \pi^{1/2} C_2 \dratdn^2}{C_3 \etastd^2}  =
\frac{\pi^2}{3 \lamfac^2} \delup^2 \epsup^{-1} \etastd^{-2} \dratdn^2 =
\frac{\pi^2}{3 \lamfac^2} \lamup^{-2} \epsup^{-3} \dratdn^2
\end{equation}
In particular, the ratio on the left, which can be measured in the experiments, is predicted to scale as the inverse cube of the wave amplitude (as well as the square of the depth ratio).

\vsp{10}
OLD VERSION: \\

To convert to dimensional, i.e.~experimental, values we use
\begin{align}
& \intt u^3 \dx = \frac{\mean{\eta^3}}{a^3} = 
\frac{\mean{\eta^3}}{\mean{\eta^2}^{3/2}} = \skw(\eta) \\
& \intt u_x^2 \dx = \left(\frac{\lambda_0}{a} \right)^2 \mean{\eta_x^2} 
= \left(\frac{\lambda_0}{a} \right)^2 \var(\eta_x)
\end{align}
Then \eqref{H3H2} gives the relationship
\begin{equation}
\frac{\skw(\eta)}{\Delta \var(\eta_x)} =
3 \mu D^2 \left( \frac{h_0}{a} \right)^3 = 3 \eps^{-3} \mu D^2
\end{equation}
In particular, the ratio on the left, which can be measured in the experiments, is predicted to scale as the inverse cube of the wave amplitude (as well as the square of the depth ratio).

>>>>

% Basic physical parameters and scales
\newcommand{\freqp}{f_p}
\newcommand{\amp}{a}
\newcommand{\ampscale}{A}
\newcommand{\etastd}{\eta_{\text std}}
% Parameters that change crossing the ADC
\newcommand{\depth}{d}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\dupdn}{\depth_{\pm}}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamdn}{\lam_{+}}
\newcommand{\lamupdn}{\lam_{\pm}}
\newcommand{\lamfac}{N}
\newcommand{\drat}{D}
\newcommand{\dratdn}{\drat_+}
\newcommand{\dratupdn}{\drat_{\pm}}
% Statistical quantities
\newcommand{\En}{\mathcal{E}}
\newcommand{\Mo}{\mathcal{M}}
\newcommand{\skw}{\text{skew}}
\newcommand{\skwdn}{\skw_+}
\newcommand{\var}{\text{var}}
\newcommand{\varup}{\var_-}
\newcommand{\vardn}{\var_+}
% Dimensionless parameters
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
% Hamiltonian and Gibbs stuff
\newcommand{\uhat}{\hat{u}}
\newcommand{\Ham}{H}
\newcommand{\Hup}{\Ham^{-}}
\newcommand{\Hdn}{\Ham^{+}}
\newcommand{\Hupdn}{\Ham^{\pm}}
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gup}{\Gibbs^{-}}
\newcommand{\Gdn}{\Gibbs^{+}}
\newcommand{\Gupdn}{\Gibbs^{\pm}}
\newcommand{\thup}{\theta^{-}}
\newcommand{\thdn}{\theta^{+}}
\newcommand{\thupdn}{\theta^{\pm}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}
\newcommand{\meanupdn}[1]{\mean{#1}_{\pm}}

@@

We will essential use the original normalization in \eqref{dimlessKdV1}, except with the slight modification that $\eta = \sqrt{\pi} \etastd u$, so that the energy of $u$ is normalized to unity. This choice will be consistent with our PNAS paper and the {\it J Stat Phys} paper of Majda and Qi.
Thus the dimensionless variables are
\begin{align}
&u = \eta/(\sqrt{\pi} \etastd)
&&\mbox{\em dimensionless surface displacement} \\
&\tilde{x} = \frac{\pi \xi}{\lamfac \lam} = \frac{\pi (x-ct)}{\lamfac \lam}
&&\mbox{\em dimensionless position in moving frame} \\
& \tilde{t} = t \freqp / \lamfac
&&\mbox{\em dimensionless time}
\end{align}
The explanation will be something like: instead of the most conventional choice, we scale displacement on $\sqrt{\pi} \etastd$ so that energy is normalize to unity. This choice will be the most convenient for the Hamiltonian framework that we will rely on. Maybe we will have to scale time differently too to get the desired powers of $\drat$, but that will make no difference for now.
Thus, the dimensionless KdV is given by
\begin{align}
\label{dimlessKdV2}
&u_t + C_3 \drat^{-1} \, u u_x + C_2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi^{3/2} \epsup  \\
&C_2 = \frac{\pi^3 \delup^2}{6 \lamfac^2} 
\end{align}
When we draw a comparison to the PNAS paper, it will be something like $\epsup = E^{1/2}L^{-3/2}$ and $\delup^2 = L^{-3}$, etc.

@@

The statistical matching condition is
\begin{equation}
\meanup{\Hdn} = \meandn{\Hdn}
\end{equation}
An experimental observation is that the incoming skewness is small, $\meanup{H_3} \approx 0$. Inserting and simplifying gives
\begin{equation}
\label{H3H2}
\meandn{H_3} = \frac{C_2}{C_3} \dratdn^2 \,\, \Delta \mean{H_2}
\end{equation}
where $\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2}$  indicates the difference between the upstream and downstream values.
To convert to dimensional, i.e.~experimental, values we note
\begin{align}
&\mean{u^3} =  \frac{\mean{\eta^3}} {\pi^{3/2} \mean{\eta^2}^{3/2}} = \pi^{-3/2} \skw(\eta) \\
&\mean{u_x^2} = \pi^{-1} \left(\frac{\lamupdn}{\etastd} \right)^2 \mean{\eta_x^2} 
= \pi^{-1} \left(\frac{\lamupdn}{\etastd} \right)^2 \var(\eta_x)
\end{align}
Therefore
\begin{align}
& \meandn{H_3} = \frac{\pi}{3} \mean{u^3} = 
\frac{1}{3} \pi^{-1/2} \skwdn(\eta) \\
& \meanupdn{H_2} = \pi \meanupdn{u_x^2} = 
\left(\frac{\lamupdn}{\etastd} \right)^2 \var(\eta_x)
\end{align}
The second equation implies
\begin{equation}
\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2} = 
\etastd^{-2} \left( \lamdn^2 \vardn(\eta_x) - \lamup^2 \varup(\eta_x) \right) =
\etastd^{-2} \Delta \left( \lam^2 \var(\eta_x) \right)
\end{equation}
Then \eqref{H3H2} gives the relationship
\begin{equation}
\frac{\skw(\eta)} {\Delta \left( \lam^2 \var(\eta_x) \right) } =
\frac{3 \pi^{1/2} C_2 \dratdn^2}{C_3 \etastd^2}  =
\frac{\pi^2}{3 \lamfac^2} \delup^2 \epsup^{-1} \etastd^{-2} \dratdn^2 =
\frac{\pi^2}{3 \lamfac^2} \lamup^{-2} \epsup^{-3} \dratdn^2
\end{equation}
Performing some additional manipulations, and noting that $(\lamdn/\lamup)^2 = \ddn/\dup = \dratdn$, gives
\begin{equation}
\frac{\skwdn(\eta)} {\dratdn \vardn(\eta_x) - \varup(\eta_x)} = 
\frac{\pi^2}{3 \lamfac^2} \epsup^{-3} \dratdn^2
\end{equation}
In particular, the ratio on the left, which can be measured in the experiments, is predicted to scale as the inverse cube of the wave amplitude and as the square of the depth ratio.

% Comparison to old calculation:
% The old calculation used mu = 4 pi^2 / (9 N^2), but with a factor of two difference in the definition of N. The resulting coefficient on the RHS was 3 mu, which agrees with the RHS above once the factor of two is taken into account.
